pdp: An R Package for Creating Partial Dependence Plots

### 2016-09-11 manuscript/greenwell.tex
minor edits

Predictor importance is an important task in any supervised learning problem. However, ranking variables is only part of the story and once a subset of "important" features is identified it is often necessary to assess the relationship between them (or subset thereof) and the response. This can be done in many ways, but in machine learning it is often accomplished by constructing \dfn{partial dependence plots} (PDPs); see \citet{friedman-2000-greedy} for details. PDPs help visualize the relationship between a subset of the predictors (typically 1-3) and the response while accounting for the average effect of the other independent variables in the model. They are particularly effective with black box models like random forest and support vector machines.
%PDPs offer a low-dimensional graphical rendering of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ for any type of fitted models and are especially useful in visualizing the relationships discovered by complex machine learning algorithms such as a \dfn{random forest}.

@@

To construct a PDP \eqref{eqn:pdf} in practice, we can proceed as follows. To simplify, let $\boldsymbol{z}_s = x_1$ be the predictor variable of interest with unique values $\left\{x_{11}, x_{12}, \dots, x_{1k}\right\}$. The partial dependence of the response on $x_1$ can be constructed as follows:
\begin{enumerate}
  \item For $i \in \left\{1, 2, \dots, k\right\}$:
  \begin{enumerate}
    \item Create a copy of the training data and replace the original values of $x_1$ with the constant $x_{1i}$.
    \item Compute the vector of predicted values from the modified copy of the training data.
    \item Compute the average prediction to obtain $\bar{f}_1\left(x_{1i}\right)$.
  \end{enumerate}
  \item Plot the pairs $\left[x_{1i}, \bar{f}_1\left(x_{1i}\right)\right]$ for $i = 1, 2, \dotsc, k$.
\end{enumerate}
This can be quite computationally intensive since the algorithm involves $k$ passes over the training records. Fortunately, this can be parallelized quite easily (more on this in a later section). This algorithm can also be easily extended to larger subsets of two or more features as well.

Limited implementations of Friedman's PDPs are available in packages \CRANpkg{randomForest} \citep{randomForest-pkg} and \CRANpkg{gbm}, among others; these are limited in the sense that they only apply to the models fit using the respective package. For example, the \code{partialPlot} function in \pkg{randomForest} only applies to objects of class \code{"randomForest"}. While the \pkg{randomForest} implementation will only allow for a single predictor, the \pkg{gbm} implementation can deal with any subset of the predictor space. Partial dependence functions are not restricted to tree-based models; they can be applied to any supervised learning algorithm (e.g., multiple linear regression and neural networks). However, to our knowledge, there is no general package for constructing PDPs  in R (e.g., PDPs for a \dfn{conditional random forest} as implemented by the \code{cforest} function in the \CRANpkg{party} and \CRANpkg{partykit} packages; see \citet{party-pkg} and \citet{partykit-pkg}, respectively). The \CRANpkg{pdp} \citep{pdp-pkg} package tries to close this gap by offering a general framework for PDPs that can be applied to several types of fitted models.

The \CRANpkg{plotmo} package \citep{plotmo-pkg} is one alternative to \pkg{pdp}. According to the author, \pkg{plotmo} constructs "a poor man's partial dependence plot." In particular, it plots a model's response when varying one or two predictors while holding the other predictors in the model constant (continuous features are fixed at their median value, while factors are held at their first level). These plots allow for up to two variables at a time and are less accurate than PDPs, but are faster to construct. For additive models (i.e., no interactions), these plots are identical in shape to PDPs.

PDPs can be misleading in the presence of substantial interactions \citep{goldstein-peeking-2015}. To overcome this issue \citeauthor*{goldstein-peeking-2015} developed the concept of \dfn{individual conditional expectation} (ICE) plots -- available in the \CRANpkg{ICEbox} package. \pkg{ICEbox} only allows for one variable at a time (i.e., no multivariate displays), though color can be used effectively to display information about an additional predictor. The ability to construct derivative plots is also available.

@@

Currently, only two functions are exported by \pkg{pdp}:
\begin{itemize}
  \item \code{partial}
  \item \code{plotPartial}
\end{itemize}
The \code{partial} function evaluates the partial dependence \eqref{eqn:pdf} from a fitted model over a grid of predictor values; the predictors are specified using the \code{pred.var} argument. If \code{plot = FALSE} (the default), \code{partial} returns an object of class \code{"partial"} which inherits from the class \code{"data.frame"}; in other words, by default, \code{partial} returns a data frame with an additional class that is recognized by the \code{plotPartial} function. The columns are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{y} and contains the values of the partial dependence function $\bar{f}_s\left(\boldsymbol{z}_s\right)$. If \code{plot = TRUE}, then \code{partial} makes an internal call to \code{plotPartial} and returns the PDP in the form of a \CRANpkg{lattice} plot \citep{lattice-pkg}. The \code{plotPartial} function can be used for displaying more advanced PDPs; it operates on objects of class \code{"partial"} and has many useful plotting options. Of course, since the default output produced by \code{partial} is still a data frame, the user can use any plotting package he/she desires to visualize the results--\CRANpkg{ggplot2} \citep{ggplot2-pkg}, for instance.

Currently supported models are described in Table~\ref{tab:models}. However, the default method \code{partial.default} is flexible enough to handle many of the model types not listed in in this table; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package. Although, for these cases, the user will need to specify the additional options \code{super.type} (i.e., \code{"regression"} or \code{"classification"}) and \code{train} (the original training data) in the call to \code{partial}---the exception in Table~\ref{tab:models} is using partial with \code{"xgb.Booster"} models, which requires the user to specify \code{train}.

@@

For a more customizable plot, call \code{partial} with \code{plot = FALSE} and use the \code{plotPartial} function. \strong{Note:} the \dfn{pipe} operator \code{\%>\%} provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is exported for writing more convenient code, as illustrated in the example below which increases the line width, adds a LOESS smoother, and customizes the $y$-axis label.
\begin{example}
boston.rf %>%
  partial(pred.var = "lstat") %>%
  plotPartial(smooth = TRUE, lwd = 2, ylab = expression(f(lstat)))
\end{example}
The result is displayed in the right side of Figure~\ref{fig:pd_lstat} below.

@@

The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of model fits (not just random forests), (2) it will allow for any number of predictors to be used, and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section. For example, the following code chunk uses a multivariate adaptive regression splines (MARS) model \citep{friedman-1991-mars} to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}. The results, which make use of various \code{plotPartial} options, are displayed in Figure~\ref{fig:pd_lstat_rm}.
\begin{example}
library(earth)  # for fitting MARS models
boston.mars <- earth(cmedv ~ ., data = boston, degree = 3)
pd <- partial(boston.mars, pred.var = c("lstat", "rm"))
plotPartial(pd)
plotPartial(pd, col.regions = colorRampPalette(c("red", "white", "blue")))
plotPartial(pd, contour = FALSE, zlab = "cmedv", drape = TRUE,
            colorkey = TRUE, screen = list(z = -20, x = -60))
\end{example}
Note that the default color map for contour plots is the Matplotlib "viridis" color map provided through the \CRANpkg{viridis} package \citep{viridis-pkg}.

@@

For classification problems, partial dependence functions are on a scale similar to the logit; see, for example, \citet[pp. 369--370]{hastie-elements-2009}. Suppose the response is categorical with $K$ levels, then for each class we compute
\begin{equation}
\label{eqn:avg-logit}
f_k(x) = \log\left[p_k(x)\right] - \frac{1}{K}\sum_{i = 1}^K\log\left[p_k(x)\right], \quad k = 1, 2, \dots, K,
\end{equation}
where $p_k(x)$ is the predicted probability for the $k$-th class. Plotting $f_k(x)$ helps us understand how the log-odds for the $k$-th class depends on different subsets of the predictor variables.

@@

XGBoost, short for eXtreme Gradient Boosting, is a popular library providing optimized distributed gradient boosting that is specifically designed to be highly efficient, flexible and portable. The associated R package \pkg{xgboost} has been used to win a number of \href{https://www.kaggle.com/}{Kaggle competitions}. It has been shown to be many times faster than the well known \pkg{gbm} package. However, unlike \pkg{gbm}, \pkg{xgboost} does not have built-in functions for displaying PDPs. Fortunately, the \pkg{pdp} package can be used to fill this gap.

For illustration, we return to the Boston housing data. The code chunk below fits an \pkg{xgboost} model to the \code{boston} data frame from Section~\ref{sec:boston}; the tuning parameters were predetermined using 5-fold cross-validation. (Make sure you are using the latest version of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.)
\begin{example}
set.seed(102)
boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
                      label = boston$cmedv, objective = "reg:linear",
                      nrounds = 2000, max_depth = 3, eta = 0.01)
\end{example}

To use \code{partial} with the \code{"xgb.Booster"} class, we need to supply the original training data (minus the response!) in the call to \code{partial}. The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together. The results are displayed in Figure~\ref{fig:boston_xgb}. Note that while \code{xgboost} requires the training data to be a matrix (or an object of class \code{"dgCMatrix"} or \code{"xgb.DMatrix"}), \code{partial} requires the training data as a data frame with factors in tact!
\begin{example}
pred.list <- list("lstat", "rm", c("lstat", "rm"))
for (x in pred.list) {
  partial(boston.xgb, pred.var = x, plot = TRUE,
          train = subset(boston, select = -cmedv))
}
\end{example}

@@

In this paper, we showed how to construct PDPs for various types of black box models in R using the \pkg{pdp} package. We also briefly discussed related approaches availble in other R packages as well. Suggestions to avoid extrpolation and high excecution times were discussed and demonstrated via examples. There is a bright future for \pkg{pdp} in terms of growth. For example, it would be useful to have the ability to construct PDPs for black box survival models---like conditional random forests with censored response. % We introduced the investr package for computing point estimates along with inversion and Waldbased confidence intervals for linear and nonlinear calibration problems with constant variance. We also showed how the boot package can be used for constructing approximate 100(1 􀀀 a)% calibration intervals, which for convenience, may be incorporated into a future release of investr. The authors are currently working on extending the package to handle the case of heteroscedastic errors, random coefficient models (e.g., objects of class "lme" from the recommended nlme package; Pinheiro et al., 2014)), and even multivariate calibration problems (e.g., objects of class "mlm").

>>>>

Predictor importance is an important task in any supervised learning problem. However, ranking variables is only part of the story and once a subset of "important" features is identified it is often necessary to assess the relationship between them (or subset thereof) and the response. This can be done in many ways, but in machine learning it is often accomplished by constructing \dfn{partial dependence plots} (PDPs); see \citet{friedman-2000-greedy} for details. PDPs help visualize the relationship between a subset of the predictors (typically 1-3) and the response while accounting for the average effect of the other independent variables in the model. They are particularly effective with black box models like random forests and support vector machines.

@@

Constructing a PDP \eqref{eqn:pdf} in practice is rather straightforward. To simplify, let $\boldsymbol{z}_s = x_1$ be the predictor variable of interest with unique values $\left\{x_{11}, x_{12}, \dots, x_{1k}\right\}$. The partial dependence of the response on $x_1$ can be constructed as follows:
\begin{enumerate}
  \item For $i \in \left\{1, 2, \dots, k\right\}$:
  \begin{enumerate}
    \item Create a copy of the training data and replace the original values of $x_1$ with the constant $x_{1i}$.
    \item Compute the vector of predicted values from the modified copy of the training data.
    \item Compute the average prediction to obtain $\bar{f}_1\left(x_{1i}\right)$.
  \end{enumerate}
  \item Plot the pairs $\left[x_{1i}, \bar{f}_1\left(x_{1i}\right)\right]$ for $i = 1, 2, \dotsc, k$.
\end{enumerate}
This can be quite computationally intensive since the algorithm involves $k$ passes over the training records. Fortunately, this can be parallelized quite easily (more on this in a later section). This algorithm can also be easily extended to larger subsets of two or more features as well.

Limited implementations of Friedman's PDPs are available in packages \CRANpkg{randomForest} \citep{randomForest-pkg} and \CRANpkg{gbm}, among others; these are limited in the sense that they only apply to the models fit using the respective package. For example, the \code{partialPlot} function in \pkg{randomForest} only applies to objects of class \code{"randomForest"}. While the \pkg{randomForest} implementation will only allow for a single predictor, the \pkg{gbm} implementation can deal with any subset of the predictor space. Partial dependence functions are not restricted to tree-based models; they can be applied to any supervised learning algorithm (e.g., multiple linear regression and neural networks). However, to our knowledge, there is no general package for constructing PDPs  in R (e.g., PDPs for a \dfn{conditional random forest} as implemented by the \code{cforest} function in the \CRANpkg{party} and \CRANpkg{partykit} packages; see \citet{party-pkg} and \citet{partykit-pkg}, respectively). The \CRANpkg{pdp} \citep{pdp-pkg} package tries to close this gap by offering a general framework for constructing PDPs that can be applied to several types of fitted models.

The \CRANpkg{plotmo} package \citep{plotmo-pkg} is one alternative to \pkg{pdp}. According to the author, \pkg{plotmo} constructs "a poor man's partial dependence plot." In particular, it plots a model's response when varying one or two predictors while holding the other predictors in the model constant (continuous features are fixed at their median value, while factors are held at their first level). These plots allow for up to two variables at a time and are less accurate than PDPs, but are faster to construct. For additive models (i.e., models with no interactions), these plots are identical in shape to PDPs.

PDPs can be misleading in the presence of substantial interactions \citep{goldstein-peeking-2015}. To overcome this issue \citeauthor*{goldstein-peeking-2015} developed the concept of \dfn{individual conditional expectation} (ICE) plots---available in the \CRANpkg{ICEbox} package. \pkg{ICEbox} only allows for one variable at a time (i.e., no multivariate displays), though color can be used effectively to display information about an additional predictor. The ability to construct derivative plots is also available.

@@

Currently, only two functions are exported by \pkg{pdp}:
\begin{itemize}
  \item \code{partial}
  \item \code{plotPartial}
\end{itemize}
The \code{partial} function evaluates the partial dependence \eqref{eqn:pdf} from a fitted model over a grid of predictor values; the predictors are specified using the \code{pred.var} argument. If \code{plot = FALSE} (the default), \code{partial} returns an object of class \code{"partial"} which inherits from the class \code{"data.frame"}; put another way, by default, \code{partial} returns a data frame with an additional class that is recognized by the \code{plotPartial} function. The columns of the data frame are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{y} and contains the values of the partial dependence function $\bar{f}_s\left(\boldsymbol{z}_s\right)$. If \code{plot = TRUE}, then \code{partial} makes an internal call to \code{plotPartial} and returns the PDP in the form of a \CRANpkg{lattice} plot \citep{lattice-pkg}. The \code{plotPartial} function can be used for displaying more advanced PDPs; it operates on objects of class \code{"partial"} and has many useful plotting options. Of course, since the default output produced by \code{partial} is still a data frame, the user can use any plotting package he/she desires to visualize the results---\CRANpkg{ggplot2} \citep{ggplot2-pkg}, for instance.

Currently supported models are described in Table~\ref{tab:models}. However, the default method \code{partial.default} is flexible enough to handle many of the model types not listed in in this table; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package. Although, for these cases, the user will need to specify the additional options \code{type} (i.e., \code{"regression"} or \code{"classification"}) and \code{train} (the original training data) in the call to \code{partial}---the exception in Table~\ref{tab:models} is using partial with the \code{"xgb.Booster"} class, which requires the user to specify \code{train}.

@@

For a more customizable plot, we can set \code{plot = FALSE} in the call to \code{partial} and then use the \code{plotPartial} function on the resulting data frame. This is illustrated in the example below which increases the line width, adds a LOESS smoother, and customizes the $y$-axis label. The result is displayed in the right side of Figure~\ref{fig:pd_lstat}. \strong{Note:} the \dfn{pipe} operator \code{\%>\%} provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is exported for writing more convenient code.
\begin{example}
boston.rf %>%
  partial(pred.var = "lstat") %>%
  plotPartial(smooth = TRUE, lwd = 2, ylab = expression(f(lstat)))
\end{example}

@@

The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of fitted models (not just random forests), (2) it will allow for any number of predictors to be used (e.g., multivariate displays), and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section. For example, the following code chunk uses a multivariate adaptive regression splines (MARS) model \citep{friedman-1991-mars} to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}. The results, which make use of various \code{plotPartial} options, are displayed in Figure~\ref{fig:pd_lstat_rm}.
\begin{example}
library(earth)  # for fitting MARS models
boston.mars <- earth(cmedv ~ ., data = boston, degree = 3)
pd <- partial(boston.mars, pred.var = c("lstat", "rm"))
plotPartial(pd)
plotPartial(pd, col.regions = colorRampPalette(c("red", "white", "blue")))
plotPartial(pd, contour = FALSE, zlab = "cmedv", drape = TRUE,
            colorkey = TRUE, screen = list(z = -20, x = -60))
\end{example}
\strong{Note:} the default color map for contour plots is the Matplotlib "viridis" color map provided through the \CRANpkg{viridis} package \citep{viridis-pkg}.

@@

For classification problems, partial dependence functions are on a scale similar to the logit; see, for example, \citet[pp. 369---370]{hastie-elements-2009}. Suppose the response is categorical with $K$ levels, then for each class we compute
\begin{equation}
\label{eqn:avg-logit}
f_k(x) = \log\left[p_k(x)\right] - \frac{1}{K}\sum_{i = 1}^K\log\left[p_k(x)\right], \quad k = 1, 2, \dots, K,
\end{equation}
where $p_k(x)$ is the predicted probability for the $k$-th class. Plotting $f_k(x)$ helps us understand how the log-odds for the $k$-th class depends on different subsets of the predictor variables.

@@

To round out our discussion, we provide one last example using a recently popular (and successful!) machine learning tool. XGBoost, short for eXtreme Gradient Boosting, is a popular library providing optimized distributed gradient boosting that is specifically designed to be highly efficient, flexible and portable. The associated R package \pkg{xgboost} has been used to win a number of \href{https://www.kaggle.com/}{Kaggle competitions}. It has been shown to be many times faster than the well known \pkg{gbm} package. However, unlike \pkg{gbm}, \pkg{xgboost} does not have built-in functions for displaying PDPs. Fortunately, the \pkg{pdp} package can be used to fill this gap.

For illustration, we return to the Boston housing data. The code chunk below fits an \pkg{xgboost} model to the \code{boston} data frame from Section~\ref{sec:boston}; the tuning parameters were predetermined using 5-fold cross-validation. (Make sure you are using the latest version of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.)
\begin{example}
set.seed(102)  # for reproducibility
boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
                      label = boston$cmedv, objective = "reg:linear",
                      nrounds = 2000, max_depth = 3, eta = 0.01)
\end{example}

To use \code{partial} with the \code{"xgb.Booster"} class, we need to supply the original training data (minus the response!) in the call to \code{partial}. The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together. The results are displayed in Figure~\ref{fig:boston_xgb}. Note that while \code{xgboost} requires the training data to be a matrix (or an object of class \code{"dgCMatrix"} or \code{"xgb.DMatrix"}), \code{partial} requires the training data as a data frame with factors in tact!
\begin{example}
pred.list <- list("lstat", "rm", c("lstat", "rm"))
for (x in pred.list) {
  partial(boston.xgb, pred.var = x, plot = TRUE, smooth = TRUE, rug = TRUE,
          chull = TRUE, train = subset(boston, select = -cmedv))
}
\end{example}

@@

In this paper, we showed how to construct PDPs for various types of black box models in R using the \pkg{pdp} package. We also briefly discussed related approaches availble in other R packages as well. Suggestions to avoid extrpolation and high excecution times were discussed and demonstrated via examples. There is a bright future for \pkg{pdp} in terms of growth. For example, it would be useful to have the ability to construct PDPs for black box survival models---like conditional random forests with censored response.
